A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-01 paper.tex
Update on parsimony, following review

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now broken,
since such parsimonious ARGs are extremely unlikely under the
% JK: I assume this is true; any citations to back it up I wonder?
coalescent with recombination.
\cite{minichiello2006mapping} made this separation of the ARG as a
stochastic process vs the ARG as a data structure explicit, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as describing algorithms for more general phylogenetic
networks~\citep{huson2010phylogenetic}.

>>>>

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
It is important to note that such recombination-parsimony ARGs
have quite different properties from realisations of the
(previously synonymous) ARG \emph{process}. Firstly, branch lengths are
not typically estimated, and node times are required
order to compute a likelihood under the coalescent with recombination
(see section XXX). Secondly, even if branch lengths were estimated,
the realisations would have a very small likelihood since
the coalescent with recombination is often highly \emph{un}parsimonious
in terms of recombination events for non-trivial scaled recombination rate.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now therefore broken.
\cite{minichiello2006mapping} made this separation of the ARG as a
stochastic process vs the ARG as a data structure explicit, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as describing algorithms for more general phylogenetic
networks~\citep{huson2010phylogenetic}.
